---
title: "Calculating Imprinting Probabilities"
author: "Zane Billings"
date: "`r Sys.Date()`"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
load_pkg <- function(pkg) {
  if (isFALSE(require(pkg, character.only = TRUE))) {
    install.packages(pkg, repos="https://cran.rstudio.com")
  } else {
    require(pkg, character.only = TRUE)
  }
}
load_pkg("tidyverse")
load_pkg("here")

# Constant for setting the maximum flu season in the data.
cyear <- 2020

# TODO check if there is an API for automatically downloading the ILINet and
#  WHO data.
```

# Introduction

In this notebook I will test how to calculate HA imprinting probabilities for H1, H2, and H3 subtypes of influenza A for each birth-year cohort and season using the method of @Arevalo2020 and @Gostic2016.

Almost everything in this document is ripped directly from these two papers and just re-written in a format that is understandable by me.

Once these calculations are completed and double-checked, they will be moved into the main CIVICdata repository and an LUT can be calculated for each season/birth-year combination, so that the imprinting probabilities can be assigned to each individual in the cohort.

The imprinting probability, $m_{s,t,y}$ is defined as the probability that a perosn born in year $y$ yhad their first influenza A infection with subtype $s$ in season $t$ is given by
$$m_{s,t,y} = l_{s,t} a_{y,t} U_{y,t}$$
where the necessary components are

* $l_{s,t}$, the frequency of subtype $s$ during season $t$;
* $a_{y,t}$, the probability that a naive individual born in year $y$ is infected in season $t$;
* ${i_{y,t}}$, the time-varying per-season infection rate;
* $i_0$, the expected per-season infection rate;
* $I_t$, the seasonal intensity of influenza A;
* $\gamma_{y,t}$, the fraction of season $t$ experienced by an individual born in year $y$; and
* $U_{y,t}$, the fraction of people born in year $y$ who were unexposed at the beginning of season $t$.

# $i_0$: expected per-season infection rate

Assuming that the rate of influenza A infection in each season is constant, say, $i_0$, we have that
$$P(\text{infection in single season}) = 1 - e^{-i_0}.$$

Assuming that the average probability of a naive individual being infected in a single season is $0.28$ [@bodewes2011, @gostic2016], then
$$i_0 = -\ln(1 - 0.28) \approx `r round(-log(1-0.28),4)`.$$

We will need this later so I have included the `R` code here even though it is trivial.

```{r i0}
i_0 <- -log(1-0.28)
```

# $I_t$: seasonal intensity

The **unnormalized** seasonal intensity, $\tilde{I_t}$, was defined as the product of $\text{ILI}_t$, the mean fraction of patients with ILI during season $t$, and $F_t$, the percentage of specimans testing positive for influenza A during season $t$; divided by the total number of respiratory specimens tested in season $t$. That is,
$$\tilde{I_t} = \frac{\text{ILI}_t F_t}{N_t}.$$

The final seasonal intensity, $I_t$, used in the calculations, was **normalized** by the mean intensity over the seasons from 1976 to `r cyear`, i.e.
$$I_t = \frac{\tilde{I_t}}{\frac{1}{`r cyear - 1976 + 1`}\sum_{k = 1976}^{`r cyear`} \tilde{I_k}}.$$

* For the seasons from 1997 through 2017:
  * $\text{ILI}_t$ data were obtained from the ILINet data available from @fluview. The data are reported as weekly (EXPLAIN HOW TO GET THE QUANTITY)
  * $F_t$ and $N_t$ data were obtained from the WHO/NREVSS data also available from @fluview. For the data from 2015 onward, week 40 of a calendar year was considered the beginning of an influenza season (so e.g. the 2015 influenza season begins in week 40 of 2015 and lasts through week 39 of 2016). Furthermore, the total number of specimens and total number of positive influenza A specimens from the clinical labs and public health surveillance were added together to obtain the totals.
* For the seasons from 1976 - 1996, seasonal ILI data were not available and it was assumed that the $\text{ILI}_t$ for these years was equal to the mean of the $\text{ILI}_t$ for the 1997 - `r cyear` seasons. The available data were obtained from @thompson2003, and the ILI data were calculated as

$$\text{ILI}_{t} = \frac{1}{`r cyear - 1997 + 1`}\sum_{k=1997}^{`r cyear`} \text{ILI}_{k}; \ t = 1976, 1977, \ldots, 1996.$$

* For all seasons prior to 1976, no seasonal intensity data were available and it was assumed that the intensity of each season was equal to the mean intensity of seasons 1976 - 2017. That is,

$$\tilde{I_t} = \frac{1}{`r cyear - 1976 + 1`}\sum_{k = 1976}^{`r cyear`} \tilde{I_k}; \ t < 1976.$$

Associated raw data files:

* `data\thompson2003.csv` contains the table from @thompson2003 with the $F_t$ and $N_t$ data used for seasons 1976 - 1996.
* `data\ILINet.csv` contains the ILINet data accessible from @fluview, used for the $\text{ILI}_t$ data from seasons 1997 - `r cyear`.
* who/nvress

In order to calculate the seasonal intensity, first we import the data used for the seasons beginning in 1976 though 1996.

```{r Thompson2003 data, warning = FALSE, message = FALSE}
thompson2003 <- 
  readr::read_csv(
    file = here::here("data/Thompson2003.csv"),
    skip = 2,
    n_max = 21,
    col_types = "cddddddd"
  ) %>%
  dplyr::transmute(
    # record season as beginning year only
    season = as.numeric(substr(Season, 0, 4)),
    # N_t is total number of specimens tested
    N_t = `Specimens\nTested`,
    # F_t is sum of H1N1 and H3N2 specimens tested
    F_t = `A(H1N1)\nPositiveIsolates` + `A(H3N2)\nPositiveIsolates`
  )
```

From 1996 onward, we can use the ILINet and WHO/NREVSS data instead, so next we import that. First we import the pre-2015 data, which is already combined.

```{r WHO data pre 2015}
who_nrevss_pre2015 <-
  readr::read_csv(
    file = here::here("data/WHO-NREVSS-pre2015.csv"),
    skip = 1,
    col_types = "ccddddddddddd"
  ) %>%
  dplyr::rowwise() %>%
  dplyr::transmute(
    year = YEAR,
    week = WEEK,
    season = ifelse(WEEK >= 40, YEAR, YEAR - 1),
    N_t = `TOTAL SPECIMENS`,
    F_t = sum(across(starts_with("A ")))
  )
```

The data post 2015 was collected from two different sources which have to be joined and summed together in order to get the totals.

```{r WHO data post 2015, message = FALSE, warning = FALSE}
who_nrevss_ph <- 
  readr::read_csv(
    file = here::here("data/WHO-NREVSS-publichealth.csv"),
    skip = 1,
    col_types = "ccdddddddddd"
  ) %>%
  dplyr::rowwise() %>%
  dplyr::transmute(
    year = YEAR,
    week = WEEK,
    season = ifelse(WEEK >= 40, YEAR, YEAR - 1),
    N_ph = `TOTAL SPECIMENS`,
    F_ph = sum(across(starts_with("A ")))
  )

who_nrevss_cl <-
  readr::read_csv(
    file = here::here("data/WHO-NREVSS-clinical.csv"),
    skip = 1,
    col_types = "ccdddddddd"
  ) %>%
  dplyr::transmute(
    year = YEAR,
    week = WEEK,
    season = ifelse(WEEK >= 40, YEAR, YEAR - 1),
    N_cl = `TOTAL SPECIMENS`,
    F_cl = `TOTAL A`
  )

who_nrevss_post2015 <-
  dplyr::full_join(
    x = who_nrevss_cl,
    y = who_nrevss_ph,
    by = c("year", "week", "season")
  ) %>%
  dplyr::rowwise(year, week, season) %>%
  dplyr::summarize(
    N_t = N_cl + N_ph,
    F_t = F_cl + F_ph,
    .groups = "drop"
  )
```

Now we just need to put the data together.

```{r WHO data}
who_nrevss <- 
  rbind(who_nrevss_pre2015, who_nrevss_post2015) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(year, week)
```

The next task for calculating seasonal intensity is to import the ILI net data so that the ILI data can be joined to this.

```{r import ILINet}
ilinet <-
  readr::read_csv(
    file = here::here("data/ILINet.csv"),
    skip = 1,
    na = c("", "NA", "X"),
    col_type = "ccddddddddddddd"
  ) %>%
  dplyr::transmute(
    week = WEEK,
    year = YEAR,
    season = ifelse(WEEK >= 40, YEAR, YEAR - 1),
    fraction_ili = `% WEIGHTED ILI` / 100,
  )
```

Now the ILI fractions need to be joined to the other data, and then $N_t$ and $F_t$ should be summed by season, and $\text{ILI}_t$ should be averaged by season.

```{r 1997 to present data joining}
intensity_post97 <- 
  dplyr::full_join(
    x = who_nrevss,
    y = ilinet,
    by = c("week", "year", "season")
  ) %>%
  dplyr::group_by(season) %>%
  dplyr::summarize(
    N_t = sum(N_t),
    F_t = sum(F_t),
    ILI_t = mean(fraction_ili),
    .groups = "drop"
  ) %>%
  dplyr::ungroup()
```

Next we can combine this with the data for the 1976 to 1996 seasons, and fill in the missing values for $\text{ILI}_t$.

```{r combining intensity components}
thompson2003_imputed <- thompson2003 %>%
  dplyr::mutate(
    ILI_t = mean(intensity_post97$ILI_t)
  )

intensity_post76 <- rbind(thompson2003_imputed, intensity_post97)
```

Now we have all the information we need to calculate seasonal intensity, and to fill in the intensity for the seasons from 1918 - 1975 as well. Finally, we need to normalize the values.

```{r seasonal intensity calculation}
seasonal_intensity <- intensity_post76 %>%
  dplyr::rowwise() %>%
  dplyr::transmute(
    season = season,
    I_t = ILI_t * F_t / N_t
  )

pre_76_intensity <- data.frame(
  season = seq(1918, 1975, 1),
  I_t = rep(mean(seasonal_intensity$I_t, na.rm = TRUE), 58)
)

# Normalize the intensity values (I-tilde_t to I_t step)
seasonal_intensity <- 
  rbind(pre_76_intensity, seasonal_intensity) %>%
  dplyr::mutate(I_t = I_t / mean(seasonal_intensity$I_t, na.rm = TRUE))
```

# $\gamma_{y,t}$: fraction of season experienced

# ${i_{y,t}}$: time-varying per-season infection rate

# $a_{y,t}$: individual time-varying probability of infection

# $U_{y,t}$: the fraction unexposed

# $l_{s,t}$: frequency of subtype $s$ during season $t$

# $m_{s,t,y}$: imprinting probability